\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-06-10 paper.qmd
fixed figure labels, cannot use _ in a label with latest versions of quarto

<!--langevitour with pD model for S-curve-->
::: {#fig-scurve_sc layout-ncol="3" fig-pos="H"}
![](figures/scurve/sc_1.png)

@@

In contrast, a poor model for the S-curve with UMAP was created using only $190$ bins and removing edges with a length greater than $0.3$. This model is inadequate as it has a higher MSE compared to the appropriate model. Visually, the model does not fit the points in some places and mismatches the pattern (@fig-scurve_sc_bad).

<!--langevitour with best pD model for S-curve-->
::: {#fig-scurve_sc_best layout-ncol="3" fig-pos="H"}
![](figures/scurve/sc_best_1.png)

@@

<!--langevitour with bad pD model for S-curve-->
::: {#fig-scurve_sc_bad layout-ncol="3" fig-pos="H"}
![](figures/scurve/sc_bad_1.png)

@@

To investigate which is the reasonable representation to visualize the five spherical Gaussian cluster data or all NLDR methods provide equivalent information, we visualize all the models in \pD{} space. Models from all NLDR methods show five well-separated clusters (see @fig-gau1_sc, @fig-gau2_sc, @fig-gau3_sc, @fig-gau4_sc, and @fig-gau5_sc). This suggests that for the five Gaussian cluster dataset, all NLDR methods effectively preserve the global structure. tSNE displays clusters with varying densities, indicating their ability to capture within-cluster variation (see @fig-gau1_sc). On the other hand, both UMAP, PHATE, PaCMAP and TriMAP show clusters with flat surfaces, suggesting a failure to capture within-cluster variation (see @fig-gau2_sc, @fig-gau3_sc, @fig-gau4_sc and @fig-gau5_sc). Therefore, UMAP, PHATE, PaCMAP and TriMAP do not capture the local structure as effectively as other methods.

::: {#fig-gau1_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_tsne_1.png)

@@

::: {#fig-gau2_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_umap_1.png)

@@

::: {#fig-gau3_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_phate_1.png)

@@

::: {#fig-gau4_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_pacmap_1.png)

@@

::: {#fig-gau5_sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_trimap_1.png)

@@

To determine whether the UMAP representation with the parameter choice suggested by @chen2023 preserves the original data structure, we visualize the model constructed with UMAP overlaid on the \pD{} data. The figures in @fig-pbmc1_sc show three well-separated clusters, indicating that the suggested UMAP representation preserves the global structure (see @fig-umap-author). However, as shown in @fig-pbmc1_sc, these clusters are close to each other in \pD{}. Also, non-linear continuity patterns and high-density patches within the clusters are observed (see @fig-pbmc1_sc). Therefore, the suggested UMAP representation (see @fig-umap-author) does not accurately preserve the local structure of the PBMC3k dataset.

@@

::: {#fig-pbmc1_sc layout-ncol="3" fig-pos="H"}
![](figures/pbmc3k/sc_1.png)

@@

We then fit the model for tSNE, and visualize the resultant model in the \pD{} data space. The model shows a quirk, as shown in @fig-pbmc2_sc. All three clusters are connected by an edge except the small and large clusters. Because the clusters are so close in \gD{}, they attempt to maintain the structure in \pD{} as well. This is evident that tSNE with perplexity $30$ provides a reasonable representation of PBMC3k data.

@@

::: {#fig-pbmc2_sc layout-ncol="3" fig-pos="H"}
![](figures/pbmc3k/sc_4.png)

@@

According to @fig-mnist1_sc1, the non-linear continuous structure observed in the \gD{} representation of PaCMAP (see @fig-pacmap-author) is also visible when visualizing the model overlaid on the data space. This indicates that PaCMAP accurately captures the structure of the \pD{} data. Additionally, the model shows a twisted pattern within the non-linear structure in \pD{} space (see @fig-mnist1_sc2), which is an additional pattern not visible in the \gD{} representation (see @fig-pacmap-author). Furthermore, as shown in @fig-mnist1_sc3, some long edges exist in the \pD{} space that are not recognized as long edges in the \gD{} representation. However, PaCMAP is a reasonable \gD{} representation of MNIST digit 1 data. Because PaCMAP preserves the local structure.

@@

::: {#fig-mnist1_sc layout-ncol="3" fig-pos="H"}
![](figures/mnist/sc_1.png){#fig-mnist1_sc1}

![](figures/mnist/sc_2.png){#fig-mnist1_sc2}

![](figures/mnist/sc_3.png){#fig-mnist1_sc3}

>>>>

<!--langevitour with pD model for S-curve-->
::: {#fig-scurve-sc layout-ncol="3" fig-pos="H"}
![](figures/scurve/sc_1.png)

@@

In contrast, a poor model for the S-curve with UMAP was created using only $190$ bins and removing edges with a length greater than $0.3$. This model is inadequate as it has a higher MSE compared to the appropriate model. Visually, the model does not fit the points in some places and mismatches the pattern (@fig-scurve-sc-bad).

<!--langevitour with best pD model for S-curve-->
::: {#fig-scurve-sc-best layout-ncol="3" fig-pos="H"}
![](figures/scurve/sc_best_1.png)

@@

<!--langevitour with bad pD model for S-curve-->
::: {#fig-scurve-sc-bad layout-ncol="3" fig-pos="H"}
![](figures/scurve/sc_bad_1.png)

@@

To investigate which is the reasonable representation to visualize the five spherical Gaussian cluster data or all NLDR methods provide equivalent information, we visualize all the models in \pD{} space. Models from all NLDR methods show five well-separated clusters (see @fig-gau1-sc, @fig-gau2-sc, @fig-gau3-sc, @fig-gau4-sc, and @fig-gau5-sc). This suggests that for the five Gaussian cluster dataset, all NLDR methods effectively preserve the global structure. tSNE displays clusters with varying densities, indicating their ability to capture within-cluster variation (see @fig-gau1-sc). On the other hand, both UMAP, PHATE, PaCMAP and TriMAP show clusters with flat surfaces, suggesting a failure to capture within-cluster variation (see @fig-gau2-sc, @fig-gau3-sc, @fig-gau4-sc and @fig-gau5-sc). Therefore, UMAP, PHATE, PaCMAP and TriMAP do not capture the local structure as effectively as other methods.

::: {#fig-gau1-sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_tsne_1.png)

@@

::: {#fig-gau2-sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_umap_1.png)

@@

::: {#fig-gau3-sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_phate_1.png)

@@

::: {#fig-gau4-sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_pacmap_1.png)

@@

::: {#fig-gau5-sc layout-ncol="3" fig-pos="H"}
![](figures/five_gau_clusters/sc_trimap_1.png)

@@

To determine whether the UMAP representation with the parameter choice suggested by @chen2023 preserves the original data structure, we visualize the model constructed with UMAP overlaid on the \pD{} data. The figures in @fig-pbmc1-sc show three well-separated clusters, indicating that the suggested UMAP representation preserves the global structure (see @fig-umap-author). However, as shown in @fig-pbmc1-sc, these clusters are close to each other in \pD{}. Also, non-linear continuity patterns and high-density patches within the clusters are observed (see @fig-pbmc1-sc). Therefore, the suggested UMAP representation (see @fig-umap-author) does not accurately preserve the local structure of the PBMC3k dataset.

@@

::: {#fig-pbmc1-sc layout-ncol="3" fig-pos="H"}
![](figures/pbmc3k/sc_1.png)

@@

We then fit the model for tSNE, and visualize the resultant model in the \pD{} data space. The model shows a quirk, as shown in @fig-pbmc2-sc. All three clusters are connected by an edge except the small and large clusters. Because the clusters are so close in \gD{}, they attempt to maintain the structure in \pD{} as well. This is evident that tSNE with perplexity $30$ provides a reasonable representation of PBMC3k data.

@@

::: {#fig-pbmc2-sc layout-ncol="3" fig-pos="H"}
![](figures/pbmc3k/sc_4.png)

@@

According to @fig-mnist1-sc1, the non-linear continuous structure observed in the \gD{} representation of PaCMAP (see @fig-pacmap-author) is also visible when visualizing the model overlaid on the data space. This indicates that PaCMAP accurately captures the structure of the \pD{} data. Additionally, the model shows a twisted pattern within the non-linear structure in \pD{} space (see @fig-mnist1-sc2), which is an additional pattern not visible in the \gD{} representation (see @fig-pacmap-author). Furthermore, as shown in @fig-mnist1-sc3, some long edges exist in the \pD{} space that are not recognized as long edges in the \gD{} representation. However, PaCMAP is a reasonable \gD{} representation of MNIST digit 1 data. Because PaCMAP preserves the local structure.

@@

::: {#fig-mnist1-sc layout-ncol="3" fig-pos="H"}
![](figures/mnist/sc_1.png){#fig-mnist1-sc1}

![](figures/mnist/sc_2.png){#fig-mnist1-sc2}

![](figures/mnist/sc_3.png){#fig-mnist1-sc3}

### 2024-06-10 paper.tex
fixed figure labels, cannot use _ in a label with latest versions of quarto

\caption{\label{fig-scurve_sc}Screen shots of the \textbf{langevitour}
of the S-curve, shows the model-in-data space, a video of the tour
animation is available at (\url{https://youtu.be/G1m4q9k--v4}).}

@@

The appropriate model for the S-curve using UMAP (n\_neighbors: \(15\))
was created with \(420\) bins. It was determined to be the best model
after removing edges with a length greater than \(0.134\). The decision
was made considering that the model accurately captures the geometry of
the S-curve by fitting the points everywhere
(Figure~\ref{fig-scurve_sc_best}).

In contrast, a poor model for the S-curve with UMAP was created using
only \(190\) bins and removing edges with a length greater than \(0.3\).
This model is inadequate as it has a higher MSE compared to the
appropriate model. Visually, the model does not fit the points in some
places and mismatches the pattern (Figure~\ref{fig-scurve_sc_bad}).

@@

\caption{\label{fig-scurve_sc_best}Screen shots of the
\textbf{langevitour} of the S-curve, shows the model-in-data space, a
video of the tour animation is available at
(\url{https://youtu.be/4RFNENX_fXk}). MSE with \(0.035\).}

@@

\caption{\label{fig-scurve_sc_bad}Screen shots of the
\textbf{langevitour} of the S-curve, shows the model-in-data space, a
video of the tour animation is available at
(\url{https://youtu.be/DHPvhkZ0lUM}). MSE with \(0.090\).}

@@

To investigate which is the reasonable representation to visualize the
five spherical Gaussian cluster data or all NLDR methods provide
equivalent information, we visualize all the models in \pD{} space.
Models from all NLDR methods show five well-separated clusters (see
Figure~\ref{fig-gau1_sc}, Figure~\ref{fig-gau2_sc},
Figure~\ref{fig-gau3_sc}, Figure~\ref{fig-gau4_sc}, and
Figure~\ref{fig-gau5_sc}). This suggests that for the five Gaussian
cluster dataset, all NLDR methods effectively preserve the global
structure. tSNE displays clusters with varying densities, indicating
their ability to capture within-cluster variation (see
Figure~\ref{fig-gau1_sc}). On the other hand, both UMAP, PHATE, PaCMAP
and TriMAP show clusters with flat surfaces, suggesting a failure to
capture within-cluster variation (see Figure~\ref{fig-gau2_sc},
Figure~\ref{fig-gau3_sc}, Figure~\ref{fig-gau4_sc} and
Figure~\ref{fig-gau5_sc}). Therefore, UMAP, PHATE, PaCMAP and TriMAP do
not capture the local structure as effectively as other methods.

@@

\caption{\label{fig-gau1_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with tSNE in \pD{},
a video of the tour animation is available at
(\url{https://youtu.be/RASEE7N5MbM}).}

@@

\caption{\label{fig-gau2_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with UMAP in \pD{},
a video of the tour animation is available at
(\url{https://youtu.be/iG4bCPkJilw}).}

@@

\caption{\label{fig-gau3_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PHATE in \pD{},
a video of the tour animation is available at
(\url{https://youtu.be/L_PVLGwfOS0}).}

@@

\caption{\label{fig-gau4_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PaCMAP in
\pD{}, a video of the tour animation is available at
(\url{https://youtu.be/z07cKXi8EJQ}).}

@@

\caption{\label{fig-gau5_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with TriMAP in
\pD{}, a video of the tour animation is available at
(\url{https://youtu.be/Chs1lYAoX2w}).}

@@

To determine whether the UMAP representation with the parameter choice
suggested by \citet{chen2023} preserves the original data structure, we
visualize the model constructed with UMAP overlaid on the \pD{} data.
The figures in Figure~\ref{fig-pbmc1_sc} show three well-separated
clusters, indicating that the suggested UMAP representation preserves
the global structure (see Figure~\ref{fig-umap-author}). However, as
shown in Figure~\ref{fig-pbmc1_sc}, these clusters are close to each
other in \pD{}. Also, non-linear continuity patterns and high-density
patches within the clusters are observed (see
Figure~\ref{fig-pbmc1_sc}). Therefore, the suggested UMAP representation
(see Figure~\ref{fig-umap-author}) does not accurately preserve the
local structure of the PBMC3k dataset.

@@

\caption{\label{fig-pbmc1_sc}Screen shots of the \textbf{langevitour} of
the PBMC3k data set, shows the model-in-data space, a video of the tour
animation is available at (\url{https://youtu.be/VqqWuE0Jj6A}).}

@@

We then fit the model for tSNE, and visualize the resultant model in the
\pD{} data space. The model shows a quirk, as shown in
Figure~\ref{fig-pbmc2_sc}. All three clusters are connected by an edge
except the small and large clusters. Because the clusters are so close
in \gD{}, they attempt to maintain the structure in \pD{} as well. This
is evident that tSNE with perplexity \(30\) provides a reasonable
representation of PBMC3k data.

@@

\caption{\label{fig-pbmc2_sc}Screen shots of the \textbf{langevitour} of
the PBMC3k data set, shows the model in high-D, a video of the tour
animation is available at (\url{https://youtu.be/5Y1hE4i7N2k}).}

@@

According to Figure~\ref{fig-mnist1_sc1}, the non-linear continuous
structure observed in the \gD{} representation of PaCMAP (see
Figure~\ref{fig-pacmap-author}) is also visible when visualizing the
model overlaid on the data space. This indicates that PaCMAP accurately
captures the structure of the \pD{} data. Additionally, the model shows
a twisted pattern within the non-linear structure in \pD{} space (see
Figure~\ref{fig-mnist1_sc2}), which is an additional pattern not visible
in the \gD{} representation (see Figure~\ref{fig-pacmap-author}).
Furthermore, as shown in Figure~\ref{fig-mnist1_sc3}, some long edges
exist in the \pD{} space that are not recognized as long edges in the
\gD{} representation. However, PaCMAP is a reasonable \gD{}
representation of MNIST digit 1 data. Because PaCMAP preserves the local
structure.

@@

\subcaption{\label{fig-mnist1_sc1}}

@@

\subcaption{\label{fig-mnist1_sc2}}

@@

\subcaption{\label{fig-mnist1_sc3}}

@@

\caption{\label{fig-mnist1_sc}Screen shots of the \textbf{langevitour}
of the MNIST digit 1 data set, shows the model-in-data space, a video of
the tour animation is available at
(\url{https://youtu.be/zcg_GXBmqjA}).}

>>>>

\caption{\label{fig-scurve-sc}Screen shots of the \textbf{langevitour}
of the S-curve, shows the model-in-data space, a video of the tour
animation is available at (\url{https://youtu.be/G1m4q9k--v4}).}

@@

The appropriate model for the S-curve using UMAP (n\_neighbors: \(15\))
was created with \(420\) bins. It was determined to be the best model
after removing edges with a length greater than \(0.134\). The decision
was made considering that the model accurately captures the geometry of
the S-curve by fitting the points everywhere
(\textbf{?@fig-scurve\_sc\_best}).

In contrast, a poor model for the S-curve with UMAP was created using
only \(190\) bins and removing edges with a length greater than \(0.3\).
This model is inadequate as it has a higher MSE compared to the
appropriate model. Visually, the model does not fit the points in some
places and mismatches the pattern (Figure~\ref{fig-scurve-sc-bad}).

@@

\caption{\label{fig-scurve-sc-best}Screen shots of the
\textbf{langevitour} of the S-curve, shows the model-in-data space, a
video of the tour animation is available at
(\url{https://youtu.be/4RFNENX_fXk}). MSE with \(0.035\).}

@@

\caption{\label{fig-scurve-sc-bad}Screen shots of the
\textbf{langevitour} of the S-curve, shows the model-in-data space, a
video of the tour animation is available at
(\url{https://youtu.be/DHPvhkZ0lUM}). MSE with \(0.090\).}

@@

To investigate which is the reasonable representation to visualize the
five spherical Gaussian cluster data or all NLDR methods provide
equivalent information, we visualize all the models in \pD{} space.
Models from all NLDR methods show five well-separated clusters (see
Figure~\ref{fig-gau1-sc}, Figure~\ref{fig-gau2-sc},
Figure~\ref{fig-gau3-sc}, Figure~\ref{fig-gau4-sc}, and
Figure~\ref{fig-gau5-sc}). This suggests that for the five Gaussian
cluster dataset, all NLDR methods effectively preserve the global
structure. tSNE displays clusters with varying densities, indicating
their ability to capture within-cluster variation (see
Figure~\ref{fig-gau1-sc}). On the other hand, both UMAP, PHATE, PaCMAP
and TriMAP show clusters with flat surfaces, suggesting a failure to
capture within-cluster variation (see Figure~\ref{fig-gau2-sc},
Figure~\ref{fig-gau3-sc}, Figure~\ref{fig-gau4-sc} and
Figure~\ref{fig-gau5-sc}). Therefore, UMAP, PHATE, PaCMAP and TriMAP do
not capture the local structure as effectively as other methods.

@@

\caption{\label{fig-gau1-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with tSNE in \pD{},
a video of the tour animation is available at
(\url{https://youtu.be/RASEE7N5MbM}).}

@@

\caption{\label{fig-gau2-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with UMAP in \pD{},
a video of the tour animation is available at
(\url{https://youtu.be/iG4bCPkJilw}).}

@@

\caption{\label{fig-gau3-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PHATE in \pD{},
a video of the tour animation is available at
(\url{https://youtu.be/L_PVLGwfOS0}).}

@@

\caption{\label{fig-gau4-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PaCMAP in
\pD{}, a video of the tour animation is available at
(\url{https://youtu.be/z07cKXi8EJQ}).}

@@

\caption{\label{fig-gau5-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with TriMAP in
\pD{}, a video of the tour animation is available at
(\url{https://youtu.be/Chs1lYAoX2w}).}

@@

To determine whether the UMAP representation with the parameter choice
suggested by \citet{chen2023} preserves the original data structure, we
visualize the model constructed with UMAP overlaid on the \pD{} data.
The figures in Figure~\ref{fig-pbmc1-sc} show three well-separated
clusters, indicating that the suggested UMAP representation preserves
the global structure (see Figure~\ref{fig-umap-author}). However, as
shown in Figure~\ref{fig-pbmc1-sc}, these clusters are close to each
other in \pD{}. Also, non-linear continuity patterns and high-density
patches within the clusters are observed (see
Figure~\ref{fig-pbmc1-sc}). Therefore, the suggested UMAP representation
(see Figure~\ref{fig-umap-author}) does not accurately preserve the
local structure of the PBMC3k dataset.

@@

\caption{\label{fig-pbmc1-sc}Screen shots of the \textbf{langevitour} of
the PBMC3k data set, shows the model-in-data space, a video of the tour
animation is available at (\url{https://youtu.be/VqqWuE0Jj6A}).}

@@

We then fit the model for tSNE, and visualize the resultant model in the
\pD{} data space. The model shows a quirk, as shown in
Figure~\ref{fig-pbmc2-sc}. All three clusters are connected by an edge
except the small and large clusters. Because the clusters are so close
in \gD{}, they attempt to maintain the structure in \pD{} as well. This
is evident that tSNE with perplexity \(30\) provides a reasonable
representation of PBMC3k data.

@@

\caption{\label{fig-pbmc2-sc}Screen shots of the \textbf{langevitour} of
the PBMC3k data set, shows the model in high-D, a video of the tour
animation is available at (\url{https://youtu.be/5Y1hE4i7N2k}).}

@@

According to Figure~\ref{fig-mnist1-sc1}, the non-linear continuous
structure observed in the \gD{} representation of PaCMAP (see
Figure~\ref{fig-pacmap-author}) is also visible when visualizing the
model overlaid on the data space. This indicates that PaCMAP accurately
captures the structure of the \pD{} data. Additionally, the model shows
a twisted pattern within the non-linear structure in \pD{} space (see
Figure~\ref{fig-mnist1-sc2}), which is an additional pattern not visible
in the \gD{} representation (see Figure~\ref{fig-pacmap-author}).
Furthermore, as shown in Figure~\ref{fig-mnist1-sc3}, some long edges
exist in the \pD{} space that are not recognized as long edges in the
\gD{} representation. However, PaCMAP is a reasonable \gD{}
representation of MNIST digit 1 data. Because PaCMAP preserves the local
structure.

@@

\subcaption{\label{fig-mnist1-sc1}}

@@

\subcaption{\label{fig-mnist1-sc2}}

@@

\subcaption{\label{fig-mnist1-sc3}}

@@

\caption{\label{fig-mnist1-sc}Screen shots of the \textbf{langevitour}
of the MNIST digit 1 data set, shows the model-in-data space, a video of
the tour animation is available at
(\url{https://youtu.be/zcg_GXBmqjA}).}
